Clearance Project - 3 - TKstats

### 2025-03-19 data-raw/ClearanceProject-3-TKstats.Rmd
Analyzed interspecies differences for fup, CLint, and httk estimates compared to in vivo data.

```{r library-loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ctxR)
library(ggdist)
library(patchwork)
```

```{r data-loading}
this_pk <- readRDS("data-raw/plasma_CvTfits.rds")
load("data-raw/chem_lists.RData")
load("data-raw/plasma_CvTfits_data.RData")
load("data-raw/rat_pars_human.RData")
all_chem_properties <- readRDS("data-raw/ctxR_3comp2_cheminfo.rds")
full_pk <- readRDS("data-raw/all_cvt_pk.rds")
full_tkstats <- get_tkstats(full_pk)
```

>>>>

```{r library-loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ctxR)
library(ggdist)
library(patchwork)
library(httk)
.onUnload <- function(libpath) {
  library.dynam.unload("httk", libpath)
}
```

```{r data-loading}
# Note that you should run ...-1-Data.Rmd before this
# RCdata/ is a directory that is not tracked by git because it contains
# large files 
this_pk <- readRDS("data-raw/plasma_CvTfits.rds")
load("data-raw/RCdata/chem_lists.RData")
load("data-raw/plasma_CvTfits_data.RData")
load("data-raw/RCdata/rat_pars_human.RData")
load("data-raw/RCdata/rat_pars_rat.RData")
all_chem_properties <- readRDS("data-raw/RCdata/ctxR_3comp2_cheminfo.rds")
full_pk <- readRDS("data-raw/all_cvt_pk.rds")
full_tkstats <- get_tkstats(full_pk)
```

## Species-specific differences in fraction unbound and intrinsic clearance

We first want to establish that there is a difference between human and rat
parameterization. Because one of our `param_df`s is only rat and the other has
substituted human parameters (scaled to a rat).

```{r rat_human_diff_fupClint}
fup_rh <- rpr_param_df %>% 
  distinct(Chemical, MW,
           Clint, Funbound.plasma) %>% 
  rename_with(\(x) {paste0(x, "_rat")},
              .cols = c(Clint, Funbound.plasma)) %>%
  left_join(
    rph_param_df %>%
      distinct(Chemical, MW,
               Clint, Funbound.plasma) %>% 
      rename_with(\(x) {paste0(x, "_human")},
                  .cols = c(Clint, Funbound.plasma))
    )

species_fup_pl <- fup_rh %>%
  ggplot(aes(
    Funbound.plasma_human,
    Funbound.plasma_rat
  )) +
  geom_abline(slope = 1) +
  geom_point() +
  geom_smooth(method = "lm", linetype = "dashed",
              color = "forestgreen") +
  labs(title = expression(f[up]),
       x = "Human", y = "Rat") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_equal(xlim = c(0,1), ylim = c(0,1), expand = TRUE)

fup_pl <- ggExtra::ggMarginal(
  species_fup_pl,
  type = "histogram",
  binwidth = 0.025,
  fill = "white"
)

lm(
  Funbound.plasma_rat ~ Funbound.plasma_human,
  data = fup_rh
)

cor.test(fup_rh$Funbound.plasma_human,
            fup_rh$Funbound.plasma_rat)

## Clint
species_clint_pl <- subset(
  fup_rh, Clint_human < 500, Clint_rat < 500
) %>%
  ggplot(aes(
    Clint_human,
    Clint_rat
  )) +
  geom_abline(slope = 1) +
  geom_abline(slope = 1, intercept = c(-50,50), linetype = "longdash") +
  geom_point() +
  # geom_smooth(method = "lm", linetype = "dashed", color = "forestgreen") +
  labs(title = expression(CL[int]),
       x = "Human", y = "Rat") +
  coord_equal(xlim = c(0,500), ylim = c(0,500)) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5,vjust = -1))

clint_pl <- ggExtra::ggMarginal(
  species_clint_pl,
  type = "histogram",
  fill = "white"
)

lm(
  Clint_rat ~ Clint_human,
  data = fup_rh
)

cor.test(fup_rh$Clint_human,
            fup_rh$Clint_rat)

# Both plots
species_fup_pl + species_clint_pl +
  plot_layout(axis_titles = "collect")

ggsave(filename = paste0(Sys.getenv("FIG_DIR"), "RestrictiveClearance_Analysis/",
                         "Rat_Human_Interspecies_FupCLint_compare.png"),
       height = 3, width = 6)

# Get the "close" estimates (+/-50)
close_clint <- fup_rh %>%
  filter(between(Clint_rat, Clint_human - 50, Clint_human + 50)) %>%
  pull(Chemical)

rph_rmsle <- rph_httk_preds %>%
  mutate(fupSpecies = "human",
         rest_sqerror = (log10(httk_Preds_restrictive) - log10(Conc))^2,
         nonrest_sqerror = (log10(httk_Preds_nonrestrictive) - log10(Conc))^2) %>%
  group_by(
    Chemical, Species, fupSpecies
  ) %>%
  summarize(rest_rmsle = sqrt(sum(rest_sqerror)/n()),
            nonrest_rmsle = sqrt(sum(nonrest_sqerror)/n())
  ) %>%
  ungroup()

rpr_rmsle <- rpr_httk_preds %>%
  mutate(fupSpecies = "rat",
        rest_sqerror = (log10(httk_Preds_restrictive) - log10(Conc))^2,
         nonrest_sqerror = (log10(httk_Preds_nonrestrictive) - log10(Conc))^2) %>%
  group_by(
    Chemical, Species, fupSpecies
  ) %>%
  summarize(rest_rmsle = sqrt(sum(rest_sqerror)/n()),
            nonrest_rmsle = sqrt(sum(nonrest_sqerror)/n())
  ) %>%
  ungroup()

bind_rows(rpr_rmsle, rph_rmsle) %>%
  mutate(close_clint = ifelse(Chemical %in% close_clint,
                              "|CLint| within 50", "|CLint| > 50")
  ) %>% 
  filter(Chemical %in% unique(rpr_rmsle$Chemical)) %>%
  ggplot(aes(nonrest_rmsle, rest_rmsle)) +
  geom_abline(slope = 1, color = "blue") +
  geom_point(aes(color = fupSpecies), shape = 21) +
  scale_color_manual(values = c("green4", "magenta3")) +
  facet_grid(cols = vars(close_clint)) +
  geom_line(aes(group = paste0(Chemical)),
            alpha = 0.4) +
  labs(x = "Nonrestrictive RMSLE", y = "Restrictive RMSLE") +
  theme_bw() +
  coord_equal()

interspecies_rmsle_trends <- bind_rows(rpr_rmsle, rph_rmsle) %>%
  mutate(
    close_clint = ifelse(Chemical %in% close_clint,
                         "|\U0394 CLint| within 50", "|\U0394 CLint| > 50")
  ) %>% 
  filter(Chemical %in% unique(rpr_rmsle$Chemical)) %>% 
  pivot_longer(cols = contains("_rmsle"),
               names_to = "clearance_type",
               values_to = "RMSLE") %>%
  pivot_wider(names_from = fupSpecies,
              values_from = RMSLE) %>%
  mutate(norm_diff = (human - rat)/rat) %>%
  filter(!is.na(norm_diff)) %>% 
  group_by(Chemical, close_clint, clearance_type) %>%
  ggplot(aes(x = forcats::fct_reorder(Chemical, norm_diff, mean),
             y = norm_diff,
             fill = clearance_type)) +
  geom_col(
    width = 0.6,
    # fill = "grey5", 
    position = position_dodge(),
    color = NA
    ) +
  labs(
    x = "Chemical",
    y = "Fold Change RMSLE\nHuman more accurate <<     >> Rat more accurate"
  ) +
  scale_fill_manual(values = c("grey20", "seagreen"),
                    breaks = c("nonrest_rmsle", "rest_rmsle"),
                    labels = c("Nonrestrictive", "Restrictive"),
                    name = "Clearance") +
  geom_hline(yintercept = 0) +
  theme_minimal() + 
  facet_grid(cols = vars(close_clint),
             # rows = vars(clearance_type),
             space = "free_x", scales = "free_x") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey5"),
        panel.background = element_rect(color = "grey5"),
        axis.title = element_text(size = 10),
        legend.position = "bottom")

species_clint_pl + interspecies_rmsle_trends +
  plot_layout(widths = c(5,8))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"), "RestrictiveClearance_Analysis/",
                         "Rat_Human_Interspecies_RMSLE_trends.png"),
       height = 5, width = 7,
       bg = "white")
```

Next what we want to do is get httk predictions for nonrestrictive & 
restrictive chemicals, but going from

```{r}
